Unshuffling Permutations

### 2015-12-13 submission.tex
Supp. annexes + ajouts demos bijection avec carres binaires

\begin{proposition} \label{prop:bijection_binary_to_permutations_squares}
    For any $n \geq 0$, the map $\BINTOPERM$ restricted to the set of
    square binary words of length $2n$ is a bijection between this last
    set and the set of square permutations of size $2n$ avoiding the
    patterns $213$ and $231$.
\end{proposition}
\begin{proof}
    [of Proposition~\ref{prop:bijection_binary_to_permutations_squares}]
    The statement of the proposition is a consequence of the
    following claims implying that $\PERMTOBIN$ is the inverse
    map of $\BINTOPERM$ over the set of square binary words.
    \begin{claim} \label{claim:binary_to_permutation_avoiding}
        The image of $\BINTOPERM$ is the set of all permutations
        avoiding $213$ and $231$.
    \end{claim}

    \begin{claim} \label{claim:square_binary_to_square_permutation}
        If $u$ is a square binary word, $\BINTOPERM(u)$ is a square
        permutation.
    \end{claim}

    \begin{claim} \label{claim:square_permutation_to_square_binary}
        If $\pi$ is a square permutation avoiding $213$ and $231$,
        $\PERMTOBIN(\pi)$ is a square binary word.
    \end{claim}
    \qed
\end{proof}

@@

\newpage
\section*{Appendix (Reviewers version only)}

\begin{proof}[of Claim~\ref{claim:binary_to_permutation_avoiding}]
    Let us first show that the image of $\BINTOPERM$ contains only
    permutations avoiding $213$ and $231$. Let $u$ be a binary word,
    $\pi = \BINTOPERM(u)$, and $P_0$ (resp. $P_1$) be the set of the
    positions of the occurrences of $0$ (resp. $1$) in $u$. By
    definition of $\BINTOPERM$, from left to right, the subword
    $v = \pi_{|P_0}$ is increasing and the subword $w = \pi_{|P_1}$
    is decreasing, and all letters of $w$ are greater than those
    of~$v$. Now, assume that $\pi$ admits an occurrence of $213$.
    Then, since $v$ is increasing and $w$ is decreasing, there is an
    occurrence of $3$ (resp. $13$, $23$) in $v$ and a relative
    occurrence of $21$ (resp. $2$, $1$). All these three cases
    contradict the fact that all letters of $w$ are greater than
    those of $v$. A similar argument shows that $\pi$ avoids~$231$
    as well.
    \smallskip

    Finally, observe that any permutation $\pi$ avoiding $213$ and
    $231$ necessarily starts by the smallest possible letter or the
    greatest possible letter. This property is then true for the
    suffix of $\pi$ obtained by deleting its first letter,
    and so on for all of its suffixes. Thus, by
    replacing each letter $a$ of $\pi$ by $0$ (resp. $1$) if $a$ has the
    role of a smallest (resp. greatest) letter, one obtains a binary
    word $u$ such that $\BINTOPERM(u) = \pi$. Hence, all permutations
    avoiding $213$ and $231$ are in the image of $\BINTOPERM$.
    \qed
\end{proof}
\bigskip

\begin{proof}[of Claim~\ref{claim:square_binary_to_square_permutation}]
    Since $u$ is a square binary word, there is a binary word $v$
    such that $u \in v \shuffle v$. Then, there are two disjoint
    sets $P$ and $Q$ of positions of letters of $u$ such that
    $u_{|P} = v = u_{|Q}$. Now, by definition of $\BINTOPERM$, the
    words $\BINTOPERM(u)_{|P}$ and $\BINTOPERM(u)_{|Q}$ have the
    same standardized $\sigma$. Hence, and by definition of
    the shuffle product of permutations, $\BINTOPERM(u)$ appears in
    $\sigma \SHUFFLE \sigma$, showing that $\BINTOPERM(u)$ is a
    square permutation.
    \qed
\end{proof}
\bigskip

\begin{proof}[of Claim~\ref{claim:square_permutation_to_square_binary}]
    Let $\pi$ be a square permutation avoiding $213$ and $231$. By
    Claim~\ref{claim:binary_to_permutation_avoiding}, $\pi$ is in
    the image of $\BINTOPERM$ and hence, $u = \PERMTOBIN(\pi)$ is a
    well-defined binary word. Since $\pi$ is a square permutation,
    there are two disjoint sets $P_1$ and $P_2$ of indexes of letters
    of $\pi$ such that $\pi_{|P_1}$ and $\pi_{|P_2}$ are
    order-isomorphic. This implies, by the definitions of $\BINTOPERM$
    and $\PERMTOBIN$, that $u_{|P_1} = u_{|P_2}$, showing that $u$
    is a square binary word.
    \qed
\end{proof}
\bigskip

\begin{proof}[of Lemma~\ref{lem:endomorphisms}]
    To prove, for $j = 1, 2, 3$, that $\phi_j$ is a morphism of
    associative algebras, we have to prove that for all permutations
    $\sigma$ and $\nu$,
    \begin{equation}
        \phi_j(\sigma \SHUFFLE \nu) =
        \phi_j(\sigma) \SHUFFLE \phi_j(\nu).
    \end{equation}
    We shall proceed here as we have done during the proof of
    Proposition~\ref{prop:shuffle_associative_commutative} by showing
    that $\phi_j$ is a morphism of coalgebras, that is
    \begin{equation}
        \Delta \phi_j = (\phi_j \otimes \phi_j) \Delta.
    \end{equation}
    In the sequel, $\pi$ is a permutation.
    \smallskip

    If $P$ is a set of indexes of letters of $\pi$, we denote by
    $\widetilde{P}$ the set $\{|\pi| - i  + 1 : i \in P\}$. Now, since
    the operation $\widetilde{\,}$ defines a bijection on the set of the
    subsets of $[|\pi|]$, and since the standardization operation
    commutes with the mirror operation on words without multiple
    occurrence of a letter, we have
    \begin{equation} \begin{split}
        \Delta(\phi_1(\pi))
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\phi_1(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_1(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\widetilde{\pi}_{|P_1}\right)
        \otimes \STD\left(\widetilde{\pi}_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\widetilde{\pi}_{|\widetilde{P_1}}\right)
        \otimes \STD\left(\widetilde{\pi}_{|\widetilde{P_2}}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \widetilde{\STD\left(\pi_{|P_1}\right)}
        \otimes \widetilde{\STD\left(\pi_{|P_2}\right)} \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \phi_1\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_1\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_1 \otimes \phi_1) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_1$ is a morphism of coalgebras and hence, that
    $\phi_1$ is a morphism of associative algebras.
    \smallskip

    Next, since by definition of the complementation operation on
    permutations, for any permutation $\tau$ and any indexes $i$ and $k$,
    we have $\tau(i) < \tau(k)$ if and only if $\bar \tau(i) > \bar \tau(k)$,
    we have
    \begin{equation} \begin{split}
        \Delta(\phi_2(\pi))
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\phi_2(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_2(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\bar \pi_{|P_1}\right)
        \otimes \STD\left(\bar \pi_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \phi_2\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_2\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_2 \otimes \phi_2) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_2$ is a morphism of coalgebras and hence, that
    $\phi_2$ is a morphism of associative algebras.
    \smallskip

    Finally, for any permutation $\tau$, if $P$ is a set of indexes of
    letters of $\tau$, we denote by $P_\tau^{-1}$ the set
    $\{\tau(i) : i \in P\}$. Since the map sending a subset $P$ of
    $[|\pi|]$ to $P_\pi^{-1}$ is a bijection, and since
    $\STD\left(\pi_{|P}\right)^{-1} = \STD\left(\pi^{-1}_{|P_\pi^{-1}}\right)$,
    we have
    \begin{equation} \begin{split}
        \Delta(\phi_3(\pi))
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\phi_3(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_3(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\pi^{-1}_{|P_1}\right)
        \otimes \STD\left(\pi^{-1}_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\pi^{-1}_{|{P_1}_\pi^{-1}}\right)
        \otimes \STD\left(\pi^{-1}_{|{P_2}_\pi^{-1}}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\pi_{|P_1}\right)^{-1}
        \otimes \STD\left(\pi_{|P_2}\right)^{-1} \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \phi_3\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_3\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_3 \otimes \phi_3) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_3$ is a morphism of coalgebras and hence, that
    $\phi_3$ is a morphism of associative algebras.
    \qed
\end{proof}
\bigskip

\begin{proof}[of Lemma~\ref{lemma:matching}]
    The proof is, naturally enough, a slight extension of
    Lemma~1 in \cite{Rizzi:Vialette:CSR:2013} and is added here in full
    for the sake of completeness.

    (\emph{1} $\Rightarrow$ \emph{2}).
    Since $\pi$ is a square, it admits a square root $\sigma$.
    Let $2n = |\pi|$ (and hence
    $|\sigma| = n$).
    By definition, there are two sets $I^1 = \{i^1_1 < i^1_2 < \dots < i^1_n\}$
    and $I^2 = \{i^2_1 < i^2_2 < \dots < i^2_n\}$ of disjoint indexes of
    $\pi$ such that $\pi_{|I^1}$ and $\pi_{|I^2}$ are both
    order-isomorphic to $\sigma$.
    Let $K_\pi$ be the linear clique associated with $\pi$ with vertices
    $\VS{K_\pi} = \{i^1_j : j \in [n]\} \cup \{i^2_j : j \in [n]\}$. It
    is easily seen that $\mathcal{M} = \{\{i^1_j, i^2_j\} : j \in [n]\}$
    is a subset of $\ES{K_\pi}$. Furthermore, $\mathcal{M}$ is a perfect
    matching since $I^1 \cap I^2 = \emptyset$, and
    $I^1 \cup I^2 = \{1, 2,  \dots 2n\}$. It is also containment-free.
    Indeed, if it is not the case then there would exist two edges
    $e = \{i^1_j, i^2_j\}$ and $e' = \{i^1_k, i^2_k\}$, $j < k$, in
    $\mathcal{M}$ such that $i^1_j < i^1_k < i^2_k < i^2_j$ or
    $i^1_j < i^2_k < i^1_k < i^2_j$. This is a contradiction since
    $i^2_k > i^2_j$ if $k > j$. Finally, for any two distinct edges
    $e = \{\i^1_j, i^2_j\}$ and $e' = \{i^1_k, i^2_k\}$ of $\mathcal{M}$,
    we have $\pi(i^1_j) < \pi(i^2_j)$ if and only if
    $\pi(i^1_k) < \pi(i^2_k)$ since we are comparing in both cases two
    elements (at position $j$ and $k$) in two patterns that are
    order-isomorphic to $\sigma$.

    (\emph{2} $\Rightarrow$ \emph{1}).
    Let $I^1 = \{i^1_j : \exists \{i_j, i_k\} \in \mathcal{M}
    \;\text{with}\; j < k\}$, and
    $I^2 = \{i^2_k : \exists \{i_j, i_k\} \in \mathcal{M}
    \;\text{with}\; j < k\}$.
    Informally, $I^1$ (resp. $I^2$) is the set of the left (resp. right)
    endpoints of the edges in $\mathcal{M}$ assuming the positions
    $1, 2, \dots, 2$ are drawn along an horizontal line.
    Without loss of generality, assume $i^1_1 < i^1_2 < \dots < i^1_n$.
    % and $I^2 = \{i^2_1, i^2_2, \dots, i^2_n\}$.
    Let $p^1 = \pi(i^1_1) \pi(i^1_2) \dots \pi(i^1_n)$ and
    $p^2 = \pi(i^2_1) \pi(i^2_2) \dots \pi(i^2_n)$ be the two patterns
    from $\pi$ induced by $I^1$ and $I^2$, respectively. Clearly $p^1$
    and $p^2$ are disjoint in $\pi$ (since $\mathcal{M}$ is a matching)
    and cover $\pi$ (since $\mathcal{M}$ is perfect). We also claim that
    $p^1$ and $p^2$ are order-isomorphic. Suppose, aiming at a
    contradiction, that this is not the case. Therefore, there exist two
    positions $j$ and $k$, say $j < k$, such that
    $\pi(i^1_j) < \pi(i^2_j)$ and $\pi(i^1_k) > \pi(i^2_k)$, or
    $\pi(i^1_j) < \pi(i^2_j)$ and $\pi(i^1_k) > \pi(i^2_k)$.
    But since $\mathcal{M}$ is containment-free, it follows that both
    $\{i^1_j, i^2_j\}$, $i^1_j < i^2_j$, and $\{i^1_k, i^2_k\}$,
    $i^1_k, i^2_k$, belong to $\mathcal{M}$. This is a contradiction
    since for any two distinct edges $(a, b)$ and $(a', b')$ in
    $\mathcal{M}$ we have $\pi(a) < \pi(a')$ if and only if
    $\pi(b) < \pi(b')$.
    \qed
\end{proof}
\bigskip

\begin{proof}[of Lemma~\ref{lemma:at most one edge monotone}]
  We first prove \emph{1}. Suppose, aiming at a contradiction, that
  there exist two $(\sigma_1, \sigma_2)$-edges $(a, a')$ and $(b, b')$
  in $\mathcal{M}$. Since $\mathcal{M}$ is containment-free, we may
  assume $a < b$ and $a' < b'$. Therefore, we have $\pi(a) < \pi(b)$
  (since $\sigma_1$ is increasing) and $\pi(a') > \pi(b')$ (since
  $\sigma_2$ is decreasing), which contradicts our hypothesis about
  $\mathcal{M}$.

  As for \emph{2}, suppose, still aiming at a contradiction, that there
  exist two $(\sigma_1, \sigma_2)$-edges $(a, a')$ and $(b, b')$ in
  $\mathcal{M}$. Since $\mathcal{M}$ is containment-free, we may assume
  $a < b$ and $a' < b'$. Therefore, we have $\pi(a) > \pi(b)$ (since
  $\sigma_1$ is decreasing) and $\pi(a') < \pi(b')$ (since $\sigma_2$ is
  increasing), which contradicts our hypothesis about~$\mathcal{M}$.
  \qed
\end{proof}

\bigskip

\begin{proof}[of Proposition~\ref{proposition:hardness}]
  Suppose first that $\sigma$ occurs in $\pi$ and fix an occurrence.
  Construct a matching $\mathcal{M}$ as follows:
  \begin{itemize}
    \item $\mathcal{M}$ contains $N_1$ pairwise crossing
    $(\nu_1, \nu'_1)$-edges.
    \item $\mathcal{M}$ contains $N_2$ pairwise crossing
    $(\nu_2, \nu'_2)$-edges.
    \item $\mathcal{M}$ contains $N_3$ pairwise crossing
    $(\nu_3, \nu'_3)$-edges.
    \item $\mathcal{M}$ contains $N_4$ pairwise crossing
    $(\nu_4, \nu'_4)$-edges.
    \item $\mathcal{M}$ contains $k+2$ pairwise crossing
    $(\sigma', \pi')$-edges as depicted in
    Figure~\ref{fig:subfig:sigma' - pi' - pi'' - sigma''}.
    More precisely,
    (i) the first position of $\sigma'$
    (\emph{i.e.}, $(2N_1+N_2+N_3) + 1$) is linked
    to the first position of $\pi'$
    (\emph{i.e.}, $(2N_1 + 2N_2 + 2N_3 + N_4 + k + 2) + 1$),
    (ii) the last position of $\sigma'$
    (\emph{i.e.}, $(2N_1+N_2+N_3) + k+2$) is linked
    to the last position of $\pi'$
    (\emph{i.e.}, $(2N_1 + 2N_2 + 2N_3 + N_4 + k + 2) + n+2$),
    and all other positions in $\sigma'$ are linked by means of $k$ pairwise
    crossing edges to the positions in
    $\pi'$ that correspond to the fixed occurence of $\sigma$ in $\pi$.
    (Notice that we use here the fact that $\sigma$ occurs in $\pi$).
    \item $\mathcal{M}$ contains $n-k$ pairwise crossing
    $(\pi', \pi'')$-edges as depicted in
    Figure~\ref{fig:subfig:sigma' - pi' - pi'' - sigma''}.
    More precisely,
    all positions in $\pi'$ that do not correspond to the fixed occurence of
    $\sigma$ in $\pi$ are linked by means of $n-k$ pairwise crossing edges
    to the positions in $\pi''$ that do not correspond to the fixed
    occurrence of $\sigma$ in $\pi$.
    \item $\mathcal{M}$ contains $k$ pairwise crossing
    $(\pi'', \sigma'')$-edges as depicted in
    Figure~\ref{fig:subfig:sigma' - pi' - pi'' - sigma''}.
    More precisely, the positions in $\pi''$ that correspond to
    the fixed occurrence of $\sigma$ in $\pi$ are linked
    by means of $k$ pairwise crossing edges to all positions in
    $\sigma''$.
    (Notice that we use here the fact that $\sigma$ occurs in $\pi$).
  \end{itemize}
  It can be easily checked (probably referring to Figure~\ref{fig:reduction}) that
  $\mathcal{M}$ is perfect, containment-free and that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
        \begin{tikzpicture}
          [
            scale=0.35,
            label/.style={anchor=base},
            cell/.style={draw,thick,fill=black!5}
          % sigma'
          \draw [cell,rounded corners] (1.5,16.5) -- (4.5,16.5) -- (4.5,19.5) -- (1.5,19.5) -- cycle;
          \draw[step=1cm,black!60,ultra thin,fill=black!10] (0.8,16.8) grid (5.2,21.2);
          \foreach \x/\y in {1/20,2/19,3/17,4/18,5/21} {
              \draw [fill=black] (\x,\y) circle (0.1);
          \node (pip) at (3,22) {$\sigma'$};
          \node[%
            draw,
            text width=5.5cm,text justified,color=black!50,rounded corners,
            rectangle callout,callout relative pointer={(0.0cm,-1cm)}
          ] at (3,26)
          {%
            $((k+1) \; \sigma \; (k+2)) \; [2N_2 + N_4 + 2n + k + 2]$
          };%
          %\node at (3,-0.5) {$((k+1) \; \sigma \; (k+2)) \; [2N_2 + N_4 + 2n + k + 2]$};
          \node at (-0.5,20) {$k+1$};
          \node at (-0.5,21) {$k+2$};
          % pi'
          \draw [cell,rounded corners] (6.5,8.5) -- (12.5,8.5) -- (12.5,14.5) -- (6.5,14.5) -- cycle;
          \draw[step=1cm,black!60,ultra thin,fill=black!10] (5.8,8.8) grid (13.2,16.2);
          \foreach \x/\y in {6/15,7/12,8/13,9/10,10/9,11/11,12/14,13/16} {
              \draw [fill=black] (\x,\y) circle (0.1);
          \node (pip) at (9.5,17) {$\pi'$};
          \node[%
            draw,
            text width=5.5cm,text justified,color=black!50,rounded corners,
            rectangle callout,callout relative pointer={(0.0cm,3.5cm)}
          ] at (9.5,-3.5)
          {%
            $((n+1) \; \pi \; (n+2)) \; [2N_2 + N_4 + n + k + 2]$
          };%
          %\node at (9.5,-2) {$((n+1) \; \pi \; (n+2)) \; [2N_2 + N_4 + n + k + 2]$};
          \node at (-0.5,15) {$n+1$};
          \node at (-0.5,16) {$n+2$};
          % pi''
          \draw [cell,rounded corners] (13.5,2.5) -- (19.5,2.5) -- (19.5,8.5) -- (13.5,8.5) -- cycle;
          \draw[step=1cm,black!60,ultra thin,fill=black!10] (13.8,2.8) grid (19.2,8.2);
          \foreach \x/\y in {14/6,15/7,16/4,17/3,18/5,19/8} {
              \draw [fill=black] (\x,\y) circle (0.1);
          \node (pipp) at (16.75,9) {$\pi''$};
          \node[%
            draw,
            text width=2.25cm,text justified,color=black!50,rounded corners,
            rectangle callout,callout relative pointer={(0.0cm,-3cm)}
          ] at (16.5,19)
          {%
            $\pi [2N_2 + N_4 + k]$
          };%
          %\node at (16.5,-3.5) {$\pi [2N_2 + N_4 + k]$};
          % sigma''
          \draw [cell,rounded corners] (19.5,-0.5) -- (22.5,-.5) -- (22.5,2.5) -- (19.5,2.5) -- cycle;
          \draw[step=1cm,black!60,ultra thin,fill=black!10] (19.8,-0.2) grid (22.2,2.2);
          \foreach \x/\y in {20/2,21/0,22/1} {
              \draw [fill=black] (\x,\y) circle (0.1);
          \node (sigmapp) at (21,3) {$\sigma''$};
          \node[%
            draw,
            text width=1.75cm,text justified,color=black!50,rounded corners,
            rectangle callout,callout relative pointer={(0.0cm,0.75cm)}
          ] at (21,-4.5)
          {%
            $\sigma [2N_2 + N_4]$
          };%
          %\node at (21,-5) {$\sigma [2N_2 + N_4]$};
          % edges sigma' - pi'
          \draw [line width=1pt]
          (1,20) .. controls +(0,9) and +(0,4) .. (6,15);
          \draw [line width=1pt]
          (5,21) .. controls +(0,9) and +(0,4) .. (13,16);
          \draw [line width=1pt]
          (2,19) .. controls +(0,-17) and +(0,-7) .. (8,13);
          \draw [line width=1pt]
          (3,17) .. controls +(0,-17) and +(0,-7) .. (9,10);
          \draw [line width=1pt]
          (4,18) .. controls +(0,-17) and +(0,-7) .. (11,11);
          % edges pi' - pi''
          \draw [line width=1pt]
          (7,12) .. controls +(0,9) and +(0,4) .. (14,6);
          \draw [line width=1pt]
          (10,9) .. controls +(0,9) and +(0,4) .. (17,3);
          \draw [line width=1pt]
          (12,14) .. controls +(0,9) and +(0,4) .. (19,8);
          % edges pi'' - sigma''
          \draw [line width=1pt]
          (15,7) .. controls +(0,-9) and +(0,-4) .. (20,2);
          \draw [line width=1pt]
          (16,4) .. controls +(0,-9) and +(0,-4) .. (21,0);
          \draw [line width=1pt]
          (18,5) .. controls +(0,-9) and +(0,-4) .. (22,1);
          % text
          % \node[%
          %   draw,
          %   text width=5.5cm,text justified,color=black!50,
          %   rectangle callout,callout relative pointer={(0.5cm,-.75cm)}
          % ] at (1,27)
          % {%
          % $2$ pairwise crossing $(\sigma',\pi')$-edges that
          % link the guards
          % };%
          % \node[%
          %   draw,
          %   text width=5.5cm,text justified,color=black!50,
          %   rectangle callout,callout relative pointer={(-0.1cm,-2.5cm)}
          % ] at (17,24)
          % {%
          % $n-k$ pairwise crossing $(\sigma',\pi')$-edges that
          % denote positions that do not correspond to
          % a specific occurence of $\sigma$ in $\pi$
          % (at positions $2$, $3$ and $5$).
          % };%
          % \node[%
          %   draw,
          %   text width=5.5cm,text justified,color=black!50,
          %   rectangle callout,callout relative pointer={(0.75cm,1.5cm)}
          % ] at (2.5,4,5)
          % {%
          % $k$ pairwise crossing $(\sigma',\pi')$-edges that
          % denote a specific occurence of $\sigma$ in $\pi$
          % (at positions $2$, $3$ and $5$).
          % };%
          % \node[%
          %   draw,
          %   text width=5.5cm,text justified,color=black!50,
          %   rectangle callout,callout relative pointer={(2cm,.75cm)}
          % ] at (7,-3)
          % {%
          % $k$ pairwise crossing $(\pi''; \sigma'')$-edges that
          % denote a specific occurence of $\sigma$ in $\pi$
          % (at positions $2$, $3$ and $5$).
          % };%
        \caption{\label{fig:subfig:sigma' - pi' - pi'' - sigma''}%
        Illustration of the containment-free perfect matching $\mathcal{M}$
        between gadgets $\sigma'$, $\pi'$, $\pi''$ and $\sigma''$ assuming two
        input permutation $\sigma = 312$ and $\pi = 4\mathbf{5}\mathbf{2}1\mathbf{3}6$
        (where a specific occurrence of $\sigma$ in $\pi$ is depicted in bold).

  Conversely, suppose that there exists a containment-free perfect matching
  $\mathcal{M}$ in $K_\mu$ such that for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
  We show that $\sigma$ occurs as a pattern in $\pi$.
  Whereas the matching $\mathcal{M}$ may not be as regular as in the forward
  direction, the main idea is to prove that $\mathcal{M}$ contains enough structure
  (more precisely, $k+2$ $(\sigma', \pi')$-edges) so that we can conclude
  that $\sigma$ occurs in $\pi$.
  We have divided the reverse direction into a serie of basic claims that
  progressively defines the overall structure of $\mathcal{M}$
  (refer to Fig.~\ref{fig:reduction}).
      \subfigure[%
        A $(\nu_1, \nu_2)$-edge $(a, b)$ together with a
        $(\nu_2, \nu'_1)$-edge $(x, y)$.
        Notice that we may have $x<b$
        (as long as $N_1 < x \leq N_1+N_2$ and $N_1 < b \leq N_1+N_2$
        do hold).
      ]{%
        \begin{tikzpicture}
          [
            scale=0.25,
            label/.style={anchor=base}
          % nu_1
          \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,9.8) grid (5.2,15.2);
          \foreach \x/\y in {0/10,1/11,2/12,3/13,4/14,5/15} {
              \draw [fill=black] (\x,\y) circle (0.1);
          % nu_4
          \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,-0.2) grid (11.2,5.2);
          \foreach \x/\y in {6/5,7/4,8/3,9/2,10/1,11/0} {
              \draw [fill=black] (\x,\y) circle (0.1);
          % nu'_1
          \draw[step=1cm,black!30,ultra thin,fill=black!10] (11.8,15.8) grid (17.2,21.2);
          \foreach \x/\y in {12/16,13/17,14/18,15/19,16/20,17/21} {
              \draw [fill=black] (\x,\y) circle (0.1);
          % labels
          \node [label] (y) at (17,14.5) {$b$};
          \node [label] (yp) at (10,-1.5) {$b'$};
          \node [label] (a) at (0,8.5) {$a$};
          \node [label] (ap) at (8,-1.5) {$a'$};
          \node (nu1) at (2.5,7.7) {$\nu_1$};
          \node (nu2) at (8.5,-2.5) {$\nu_2$};
          \node (nu1') at (14.5,13.5) {$\nu'_1$};
          % crossings edges
          \draw [line width=1pt]
          (1,11) .. controls +(0,12) and +(0,4) .. (12,16);
          \draw [line width=1pt]
          (2,12) .. controls +(0,12) and +(0,4) .. (13,17);
          \draw [line width=1pt]
          (3,13) .. controls +(0,12) and +(0,4) .. (14,18);
          \draw [line width=1pt]
          (4,14) .. controls +(0,12) and +(0,4) .. (15,19);
          \draw [line width=1pt]
          (5,15) .. controls +(0,12) and +(0,4) .. (16,20);
          % data edges
          \draw [line width=1pt]
          (0,10) .. controls +(0,8) and +(0,20) .. (8,3);
          \draw [line width=1pt]
          (10,1) .. controls +(0,32) and +(0,10) .. (17,21);
        \label{subfig:no (nu_1, nu_2)-edge - 1}
      }% end subfigure
      \qquad
      \subfigure[%
        A $(\nu_1, \nu'_1)$-edge $(x', y')$ together with an
        edge $(y, z)$ which is neither
        a $(\nu_1, \nu'_1)$-edge,
        nor a $(\nu_2, \nu'_1)$-edge,
        nor a $(\nu'_1, \nu'_1)$-edge.
        Notice that $y$ may be any position satisfying
        $N_1+N_2 < y \leq 2N_1+N_2$.
      ]{%
      \begin{tikzpicture}
        [
          scale=0.25,
          label/.style={anchor=base}
        % nu_1
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,9.8) grid (5.2,15.2);
        \foreach \x/\y in {0/10,1/11,2/12,3/13,4/14,5/15} {
            \draw [fill=black] (\x,\y) circle (0.1);
        % nu_4
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,-0.2) grid (11.2,5.2);
        \foreach \x/\y in {6/5,7/4,8/3,9/2,10/1,11/0} {
            \draw [fill=black] (\x,\y) circle (0.1);
        % nu'_1
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (11.8,15.8) grid (17.2,21.2);
        \foreach \x/\y in {12/16,13/17,14/18,15/19,16/20,17/21} {
            \draw [fill=black] (\x,\y) circle (0.1);
        % labels
        \node [label] (y) at (17,14.5) {$b$};
        \node [label] (ypp) at (23,-4.5) {$b''$};
        \node [label] (a) at (0,8.5) {$a$};
        \node [label] (a) at (2,8.5) {$x$};
        \node [label] (a) at (14,14.5) {$x'$};
        \node [label] (ap) at (8,-1.5) {$a'$};
        \node (nu1) at (2.5,7.7) {$\nu_1$};
        \node (nu2) at (8.5,-2.5) {$\nu_2$};
        \node (nu1') at (14.5,13.5) {$\nu'_1$};
        % crossings edges
        \draw [line width=1pt]
        (1,11) .. controls +(0,12) and +(0,4) .. (12,16);
        \draw [line width=1pt]

>>>>

\begin{proposition} \label{prop:bijection_binary_to_permutations_squares}
    For any $n \geq 0$, the map $\BINTOPERM$ restricted to the set of
    square binary words of length $2n$ is a bijection between this last
    set and the set of square permutations of size $2n$ avoiding the
    patterns $213$ and $231$.
\end{proposition}
\begin{proof}
    [of Proposition~\ref{prop:bijection_binary_to_permutations_squares}]
    The statement of the proposition is a consequence of the
    following claims implying that $\PERMTOBIN$ is the inverse
    map of $\BINTOPERM$ over the set of square binary words.
    \begin{claim} \label{claim:binary_to_permutation_avoiding}
        The image of $\BINTOPERM$ is the set of all permutations
        avoiding $213$ and $231$.
    \end{claim}
    \begin{proof}[of Claim~\ref{claim:binary_to_permutation_avoiding}]
        Let us first show that the image of $\BINTOPERM$ contains only
        permutations avoiding $213$ and $231$. Let $u$ be a binary word,
        $\pi = \BINTOPERM(u)$, and $P_0$ (resp. $P_1$) be the set of the
        positions of the occurrences of $0$ (resp. $1$) in $u$. By
        definition of $\BINTOPERM$, from left to right, the subword
        $v = \pi_{|P_0}$ is increasing and the subword $w = \pi_{|P_1}$
        is decreasing, and all letters of $w$ are greater than those
        of~$v$. Now, assume that $\pi$ admits an occurrence of $213$.
        Then, since $v$ is increasing and $w$ is decreasing, there is an
        occurrence of $3$ (resp. $13$, $23$) in $v$ and a relative
        occurrence of $21$ (resp. $2$, $1$). All these three cases
        contradict the fact that all letters of $w$ are greater than
        those of $v$. A similar argument shows that $\pi$ avoids~$231$
        as well.
        \smallskip

        Finally, observe that any permutation $\pi$ avoiding $213$ and
        $231$ necessarily starts by the smallest possible letter or the
        greatest possible letter. This property is then true for the
        suffix of $\pi$ obtained by deleting its first letter,
        and so on for all of its suffixes. Thus, by
        replacing each letter $a$ of $\pi$ by $0$ (resp. $1$) if $a$ has the
        role of a smallest (resp. greatest) letter, one obtains a binary
        word $u$ such that $\BINTOPERM(u) = \pi$. Hence, all permutations
        avoiding $213$ and $231$ are in the image of $\BINTOPERM$.
        \qed
    \end{proof}

    \begin{claim} \label{claim:square_binary_to_square_permutation}
        If $u$ is a square binary word, $\BINTOPERM(u)$ is a square
        permutation.
    \end{claim}
    \begin{proof}[of Claim~\ref{claim:square_binary_to_square_permutation}]
        Since $u$ is a square binary word, there is a binary word $v$
        such that $u \in v \shuffle v$. Then, there are two disjoint
        sets $P$ and $Q$ of positions of letters of $u$ such that
        $u_{|P} = v = u_{|Q}$. Now, by definition of $\BINTOPERM$, the
        words $\BINTOPERM(u)_{|P}$ and $\BINTOPERM(u)_{|Q}$ have the
        same standardized $\sigma$. Hence, and by definition of
        the shuffle product of permutations, $\BINTOPERM(u)$ appears in
        $\sigma \SHUFFLE \sigma$, showing that $\BINTOPERM(u)$ is a
        square permutation.
        \qed
    \end{proof}

    \begin{claim} \label{claim:square_permutation_to_square_binary}
        If $\pi$ is a square permutation avoiding $213$ and $231$,
        $\PERMTOBIN(\pi)$ is a square binary word.
    \end{claim}
    \begin{proof}[of Claim~\ref{claim:square_permutation_to_square_binary}]
        Let $\pi$ be a square permutation avoiding $213$ and $231$. By
        Claim~\ref{claim:binary_to_permutation_avoiding}, $\pi$ is in
        the image of $\BINTOPERM$ and hence, $u = \PERMTOBIN(\pi)$ is a
        well-defined binary word. Since $\pi$ is a square permutation,
        there are two disjoint sets $P_1$ and $P_2$ of indexes of letters
        of $\pi$ such that $\pi_{|P_1}$ and $\pi_{|P_2}$ are
        order-isomorphic. This implies, by the definitions of $\BINTOPERM$
        and $\PERMTOBIN$, that $u_{|P_1} = u_{|P_2}$, showing that $u$
        is a square binary word.
        \qed
    \end{proof}
    \qed
\end{proof}

@@
